Calibrating seed-based alignment heuristics with Sesame

### 2020-03-11 manuscript.tex
About to simplify the skip transfer matrix

It will be useful to consider a more general problem where we
have two sequences of interest labelled $+$ and $-$. The $+$ sequence
stands for the target and that the $-$ sequence stands for its duplicate.
We then define the dual alphabet $\tilde{\mathcal{A}}_0 = \{\square, |,
\downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/i}^-$
($i \geq 1$) signify that the nucleotide is a mismatch against the $-$
sequence only, the symbols $\downarrow_{/i}^+$ ($i \geq 1$) signify that
it is a mismatch against the $+$ sequence only, and the symbol
$\Downarrow$ signifies that it is a mismatch against both. As before,
every other nucleotide is replaced by the symbol $\square$, and the
terminator $|$ is appended to the end of the read. We again define reads
as sequences of segments, except that now the terminators are the symbols
$\downarrow_{/i}^-$, $\downarrow_{/i}^+$ and $\Downarrow$ ($i \geq 1$).
The tail, as usual, is terminated by the symbol $|$.

The index $i$ in the symbol $\downarrow_{/i}^-$ indicates the match
length of the $+$ sequence. Likewise, the index $i$ in the symbol
$\downarrow_{/i}^+$ indicates the match length of the $-$ sequence. For
instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is a
mismatch against the $-$ sequence, that it is a match for the $+$
sequence, and that the six preceding nucleotides were also a match for the
$+$ sequence (but the nucleotide before that was a mismatch against the
$+$ sequence). The terminators thus encode the local state of the read.

Figure~\ref{fig:dual} shows an example of read in the dual encoding. The
$+$ and $-$ sequences are shown below the read, with matches represented
as open squares and mismatches as closed squares. It is visible from this
example that symbols $\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ alternate
whenever the mismatches hit different sequences. The symbol $\Downarrow$
occurs only when a nucleotide is a double mismatch.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_dual.pdf}
\caption{\textbf{Example of dual encoding.}
An example of read is represented in the dual alphabet. The symbols
$\downarrow_{/i}^-$ represent a mismatch against the $-$ sequence, the
symbols $\downarrow_{/i}^+$ represent a mismatch against the $+$ sequence,
and the symbol $\Downarrow$ represent a mismatch against both. The index
$i$ is the match length of the sequence that is not mismatched. The
symbolic $+$ and $-$ sequences are represented below, where an open square
stands for a match and a closed square stands for a mismatch.}
\label{fig:dual}
\end{figure}

Let us assume that for each nucleotide, $a$ is the probability that the
read matches both sequences, $b$ is the probability that it matches only
the $+$ sequence, $c$ is the probability that it matches only the $-$
sequence and $d$ is the probability that it matches neither. Obviously
$a+b+c+d=1$. These definitions allow us to specify the weighted generating
function of the symbols and of the segments. To specify the transfer
matrix, we only have to make sure that we eliminate the combinations that
create a match of size $\gamma$ or more for any of the two sequences. For
notational convenience, we define
\begin{gathered}
r_i^+(z) = (az)^i cz, \\
r_i^-(z) = (az)^i bz, \\
R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
F_i(z) = 1 + az + \ldots + (az)^i,
\end{gathered}
and we can verify that the expression of the transfer matrix
$\tilde{M}_0(z)$ is

@@

As before, the term of interest is the top right entry of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
every read can be prepended by $\Downarrow$ segments and only by those
(every other terminator would imply that one of the two sequences has a
nonzero match size at the start of the read). Thus reads are precisely the
sequences of segments that can follow the symbol $\Downarrow$ and that are
terminated by a tail, the weighted generating function of which is located
in the top right entry of the matrix.

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is actually easier to
proceed as in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, we define $\tilde{p}$
as the upper bound on the probability of a mismatch, \textit{i.e.}
$\tilde{p} = \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now
gives an upper bound of the probability that a read of size $k$ contains
$s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

We illustrate the strategy delineated above for reads of size $k=50$
sequenced with an instrument with error rate $p=0.01$, when using exact
seeds of size $\gamma=17$.

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off target seeding probabilities (exact seeds).}
The curves show the probability that seeding is off target for exact seeds
of size $\gamma=17$ in reads of size $k=50$ nucleotides sequenced with an
error rate $p=0.01$. Each line shows a different number of duplicate
sequences $N$ from 1 to 10 and the x-axis shows the divergence rate $\mu$,
defined as the probability that a given duplicate differs from the target
at any given position.}
\label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is clearly visible from expression (\ref{eq:Poff}). This
can also be understood intuitively because the probability of null seeding
decreases when there are more potential candidates, and since the
probability of on-target seeding does not change, the probability of
off-target seeding must increase.

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.08. When $\mu$ is much smaller, the sequences of the
duplicates are close to that of the target so it is unlikely that the
candidate set contains one but not the other. When $\mu$ is much larger,
the sequences of the duplicates are far from that of the target and they
are unlikely to be in the candidate set. In expression (\ref{eq:Poff}),
the only term that depends on $\mu$ is $P(\overline{S}_0 \cap
\overline{S}_1)$, and it is obvious that the minimum of expression
(\ref{eq:Poff}) corresponds to the maximum of $P(\overline{S}_0 \cap
\overline{S}_1)$. This is why the worst value of $\mu$ is the same for
every $N$.

For comparison purposes, it is important to notice the order of magnitude
of the probabilities. Up to $N=3$ duplicates, off-target seeding has
probability lower than $10^{-4}$ and for $N=10$ duplicates the upper bound
is $3\cdot10^{-4}$. This sets a baseline to compare with more elaborate
seeding methods.

@@

To compute the probability that the seeding process is off target when
using skip seeds, we observe that the logic of section~\ref{sec:offexact}
can be transposed with few modifications. In particular, the probability
can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
event that the read has a \emph{skip} seed for the target (instead of an
exact seed) and $S_1$ is the event that it has a \emph{skip} seed for the
first duplicate (instead of an exact seed).

@@

Once again, we start by recoding the reads in a specialized alphabet. We
define the skip-$n$ dual alphabet $\tilde{\mathcal{A}}_n = \{\square, |,
\Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
\downarrow^-_{/1}, \downarrow^-_{/2}, \ldots, \downarrow^+_{/1},
\downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
$\downarrow^-_{/i}$ and $\downarrow^+_{/i}$ ($i \geq 1$) have the same
meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
$\Downarrow_i$ ($0 \leq i \leq n$) indicate that both sequences have match
length 0 and that the next non-skipped position is $i$ nucleotides
downstream.

Figure~\ref{fig:skip_dual} shows the read from Figure~\ref{fig:dual}
represented in the skip-3 dual encoding. It is important to note several
differences with Figure~\ref{fig:dual}. The first is that the symbols
$\Downarrow_i$ are not always associated with double mismatches. For
instance, the symbol $\Downarrow_0$ on the right side of the read
corresponds to a mismatch for the $+$ sequence only. This happens whenever
the $+$ and the $-$ sequences are mismatched in the same interval between
non-skipped positions (the mismatches do not need to be on the same
nucleotide).

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip_dual.pdf}
\caption{\textbf{Example skip dual encoding.}
The read of Figure~\ref{fig:dual} is represented in the skip dual alphabet.
The vertical bars above the read indicate non-skipped positions. The
symbols $\downarrow_{/i}^-$ and $\downarrow_{/i}^+$ have the same meaning
as in the dual alphabet. The symbols $\Downarrow_i$ indicate that both
sequences have match length 0 and that the next non-skipped position is
located $i$ nucleotides downstream. Other features are as in
Figure~\ref{fig:dual}.}
\label{fig:skip_dual}
\end{figure}

@@

We illustrate the strategy delineated above Using the same settings as in
section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=17$), except that
we use skip-9 seeds instead of exact seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_skip.pdf}
\caption{\textbf{Off target seeding probabilities (skip seeds).}
The curves show the probability that seeding is off target for skip-9
seeds of size $\gamma=17$. Read size and error rate are the same as in
Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each line
shows a different number of duplicate sequences $N$ from 1 to 10 and the
x-axis shows the divergence rate $\mu$, defined as the probability that a
given duplicate differs from the target at any given position.}
\label{fig:curves_skip}
\end{figure}

@@

We illustrate the strategy delineated above 
using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
and $\gamma=17$), except that we replace exact seed by MEM seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_mem.pdf}
\caption{\textbf{Off target seeding probabilities (MEM seeds).}
The curves show the probability that seeding is off target for MEM
seeds of minimum size $\gamma=17$. Read size and error rate are the same
as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. Each
line shows a different number of duplicate sequences $N$ from 1 to 10 and
the x-axis shows the divergence rate $\mu$, defined as the probability
that a given duplicate differs from the target at any given position.}
\label{fig:curves_mem}
\end{figure}

@@

The value of $\beta$ has little importance if $p$ is small. Say that
$p=0.01$ and that a read generates a hit in the genome such that 33
nucleotides need to be aligned after seeding. If the hit is random there
is a 99.99\% chance that at least 15 are mismatches. The denominator of
expression (\ref{eq:bayes}) is approximately $1 +
4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
support for the hypothesis that the hit is random is overwhelming.
Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
fewer nucleotides are mismathes. In this case, the denominator is
approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
10^{-9}$, the support for the hypothesis that the hit is not random is
overwhelming. If $p$ is small, we do not need to worry about the value of
$\beta$, one can choose for instance $\beta=1/2$ so that the term
$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).

Since the probability that the best candidate is a random sequence is
either very small or very large, it has no influence in the first case,
or it dominates the probability of a false positive in the secondi case.
For every read, we can use as confidence score the maximum of this
probability and of the estimated probability that seeding is off target.

>>>>

It will be useful to consider a more general problem where we
have two sequences of interest labelled $+$ and $-$. The $+$ sequence
stands for the target and that the $-$ sequence stands for its duplicate.
We then define the dual alphabet $\tilde{\mathcal{A}}_0 = \{\square, |,
\downarrow_{/1}^-, \downarrow_{/2}^-, \ldots, \downarrow_{/1}^+,
\downarrow_{/2}^+, \ldots, \Downarrow\}$. The symbols $\downarrow_{/j}^-$
($j \geq 1$) signify that the nucleotide is a mismatch against the $-$
sequence only, the symbols $\downarrow_{/j}^+$ ($j \geq 1$) signify that
it is a mismatch against the $+$ sequence only, and the symbol
$\Downarrow$ signifies that it is a mismatch against both. As before,
every other nucleotide is replaced by the symbol $\square$, and the
terminator $|$ is appended to the end of the read. We again define reads
as sequences of segments (zero or more $\square$ symbols followed by a
terminator), except that now the terminators are the symbols
$\downarrow_{/j}^-$, $\downarrow_{/j}^+$ and $\Downarrow$ ($j \geq 1$).
The tail, as usual, is terminated by the symbol $|$.

The index $j$ in the symbol $\downarrow_{/j}^-$ indicates the match length
of the $+$ sequence (note that this is not the same as the number of
$\square$ symbols in the segment). Likewise, the index $j$ in the symbol
$\downarrow_{/j}^+$ indicates the match length of the $-$ sequence. For
instance, the symbol $\downarrow_{/7}^-$ indicates that the nucleotide is
a mismatch against the $-$ sequence, that it is a match for the $+$
sequence, and that the six preceding nucleotides were also a match for the
$+$ sequence (but the nucleotide before that was a mismatch against the
$+$ sequence). The terminators thus encode the local state of the read.

Figure~\ref{fig:dual} shows an example of read in the dual encoding. The
$+$ and $-$ sequences are shown below the read, with matches represented
as open squares and mismatches as closed squares. It is visible from this
example that symbols $\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ alternate
whenever the mismatches hit different sequences. The symbol $\Downarrow$
occurs only when a nucleotide is a double mismatch.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_dual.pdf}
\caption{\textbf{Example of dual encoding.}
An example of read is represented in the dual alphabet. The symbols
$\downarrow_{/j}^-$ ($j \geq 1$) represent a mismatch against the $-$
sequence, the symbols $\downarrow_{/j}^+$ ($i \geq 1$) represent a
mismatch against the $+$ sequence, and the symbol $\Downarrow$ represents
a mismatch against both. The index $i$ is the match length of the sequence
that is not mismatched. The symbolic $+$ and $-$ sequences are represented
below, where an open square stands for a match and a closed square stands
for a mismatch.}
\label{fig:dual}
\end{figure}

Let us assume that for each nucleotide, $a$ is the probability that the
read matches both sequences, $b$ is the probability that it matches only
the $+$ sequence, $c$ is the probability that it matches only the $-$
sequence and $d$ is the probability that it matches neither. Since there
are no other cases, we have $a+b+c+d=1$.

With these definitions, $a$, $b$, $c$ and $d$ are the probabilities of
occurrence of the symbols $\square$, $\downarrow_{/j}^-$,
$\downarrow_{/j}^+$ and $\Downarrow$, respectively ($j \geq 1$). The next
sections clarify how this is used to compute the weighted gerating
functions of interest.

\subsection{Segments following $\Downarrow$}

After a $\Downarrow$ terminator, the match counter for both sequences is
reset; the following segment can thus have up to $\gamma-1$ matches for
any of the two sequences. Each match corresponds to the $\square$ symbol
with weighted generating function $az$. The terminators $\Downarrow$ and
$|$ have respective generating function $dz$ and $1$ (recall that the tail
symbol has size 0), so if the next terminator is $\Downarrow$ or $|$, the
segments have weighted generating functions $(1 + az + \ldots +
(az)^{\gamma-1})dz$ or $1 + az + \ldots + (az)^{\gamma-1}$, respectively.

If the next terminator is $\downarrow_{/i}^+$, there is a match of size
$i$ for the $-$ sequence, so the segment contains $i-1$ symbols
$\square$ (one of the matches is accounted for by the terminator). The
weighted generating function is thus $(az)^{i-1}cz$. Likewise, if the next
terminator is $\downarrow_{/i}^-$, the weighted generating function of the
segment is $(az)^{i-1}bz$. The terminators $\downarrow_{/i}^+$ and
$\downarrow_{/i}^-$ are disallowed for $j \geq \gamma$ because this would
create a match of size $\gamma$ or greater for one of the sequences.

\subsection{Segments following $\downarrow_{/j}^+$}

At a $\downarrow_{/j}^+$ terminator, the match length of the $+$ sequence
is 0 and the match length of the $-$ sequence is $j$. The next segment can
thus have up to $\gamma-1$ matches for the $+$ sequence, but only
$\gamma-j-1$ matches for the $-$ sequence, lowering the maximum size of
the segment. If the next terminator is $\Downarrow$ or $|$, the segments
have weighted generating functions $(1 + az + \ldots +
(az)^{\gamma-j-1})dz$ and $1 + az + \ldots + (az)^{\gamma-j-1}$,
respectively.

If the next terminator is $\downarrow_{/i}^-$, the weighted generating
function is $(az)^{i-1}bz$ as in the previous section. The difference is
that the terminators $\downarrow_{/i}^-$ are allowed only for $1 \leq i
\leq \gamma-j$, otherwise this would create a match of size $\gamma$ or
greater for the $-$ sequence.

If the next terminator is $\downarrow_{/i}^+$, the situation is slightly
more complicated, because this imposes $j < i \leq \gamma-1$. Indeed,
there were already $j$ matches for the $-$ sequence at the terminator
$\downarrow_{/j}^+$, and there will be more at the end of the segment
because it has no mismatch for the $-$ sequence. Taking this into account,
we see that the weighted generating function of those segments is
$(az)^{i-j-1}cz$.

\subsection{Segments following $\downarrow_{/j}^-$}

We can find the weighted generating functions by just reversing the $+$
and $-$ signs in the previous section. This way, we see that the weighted
generating function of the segments terminated by $\Downarrow$ or $|$ are
$(1 + az + \ldots + (az)^{\gamma-j-1})dz$ and $1 + az + \ldots +
(az)^{\gamma-j-1}$, respectively.

Likewise, the weighted generating function of the segments terminated by
$\downarrow_{/i}^+$ is $(az)^{i-1}cz$, where $1 \leq i \leq \gamma-j$; and
the weighted generating function of the segments terminated by
$\downarrow_{/i}^-$ is $(az)^{i-j-1}bz$, where $j < i \leq \gamma-1$.

\subsection{Transfer matrix}

We now have all the elements to specify the transfer matrix or reads with
no seed for either sequence. Recall that $a$ is the probability of a
double match, $b$ is the probability of a mismatch only against the $-$
sequence, $c$ is the probability of a mismatch only against the $+$
sequence and $d$ is the probability of a double mismatch. For notational
convenience, we define
\begin{equation}
% These expressions are not needed for reference in the text, but
% they are still numbered so that the code can reference to it.
\begin{gathered}
r_i^+(z) = (az)^i cz, \\
r_i^-(z) = (az)^i bz, \\
R_i(z) = \big( 1 + az + \ldots + (az)^i \big) dz, \\
F_i(z) = 1 + az + \ldots + (az)^i.
\end{gathered}
\end{equation}

With these notations, the information from the previous sections can be
summarized in the transfer matrix $\tilde{M}_0(z)$ equal to

@@

As before, the term of interest is the top right entry of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. To see why, observe that
every read can be prepended by $\Downarrow$-segments and only by those
(every other terminator would imply that one of the two sequences has a
nonzero match size at the start of the read). Thus reads are precisely the
sequences of segments that can follow the symbol $\Downarrow$ and that are
terminated by a tail, the weighted generating function of which is located
in the top right entry of the matrix.

$\tilde{M}_0(z)$ is too complex to compute a closed expression of
$\tilde{M}_0(z)\cdot(I-\tilde{M}_0(z))^{-1}$. It is easier to proceed as
in section~\ref{sec:example_skip} and to compute the powers of
$\tilde{M}_0(z)$ up to a finite value. This is done once again using the
arithmetic of truncated polynomials. Since each segment except the tail
contains a mismatch against at least one sequence, we define $\tilde{p}$
as the upper bound on the probability of a mismatch, \textit{i.e.}
$\tilde{p} = \max\{b,c,d\}$. The updated formula (\ref{eq:bound}) now
gives an upper bound of the probability that a read of size $k$ contains
$s$ or more mismatches as
\begin{equation*}
Pr(X \geq s) \leq \exp \left( (s-k)\log \frac{k-s}{k(1-\tilde{p})} -s\log
\frac{s}{k\tilde{p}} \right).
\end{equation*}

@@

We illustrate the strategy delineated above for reads of size $k=50$
sequenced with an instrument with error rate $p=0.01$, when using exact
seeds of size $\gamma=19$.

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off-target seeding probabilities (exact seeds).}
The surfaces show the probability that seeding is off target for exact seeds
of size $\gamma=19$ in reads of size $k=50$ nucleotides sequenced with an
error rate $p=0.01$. The divergence rate $\mu$, defined as the probability
that a given duplicate differs from the target at any given position. Note
the factor 1000 in the scale of the z-axis.}
\label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is also clearly from expression (\ref{eq:Poff}). This can
also be understood intuitively because the probability of null seeding
decreases when there are more potential candidates, and since the
probability of on-target seeding does not change, the probability of
off-target seeding must increase.

The second observation is that there exists a ``worst'' value of $\mu$
situated around 0.07. When $\mu$ is much smaller, the duplicates tend to
be exactly identical to the target, making it impossible that there is a
seed for a duplicate but no seed for the target. When $\mu$ is much
larger, the duplicates are far from the target and they are unlikely to be
in the candidate set. In expression (\ref{eq:Poff}), the only term that
depends on $\mu$ is $P(\overline{S}_0 \cap \overline{S}_1)$, and it is
clear that the minimum of expression (\ref{eq:Poff}) corresponds to the
maximum of $P(\overline{S}_0 \cap \overline{S}_1)$. This is why the worst
value of $\mu$ is the same for every $N$.

@@

To compute the probability that the seeding process is off target when
using skip seeds, we observe that the logic of section~\ref{sec:offexact}
can be transposed with few modifications. In particular, the probability
can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
event that the read has a \emph{skip} seed for the target (instead of an
exact seed) and $S_1$ is the event that the read has a \emph{skip} seed
for the first duplicate (instead of an exact seed).

@@

Again we recode the reads in a specialized alphabet. We define the
skip-$n$ dual alphabet as $\tilde{\mathcal{A}}_n = \{\square, *, |,
\Downarrow_0, \Downarrow_1, \Downarrow_2, \ldots, \Downarrow_n,
\downarrow^-_{/1}, \downarrow^-_{/2}, \ldots, \downarrow^+_{/1},
\downarrow^+_{/2}, \ldots\}$. The symbols $\square$, $|$,
$\downarrow^-_{/j}$ and $\downarrow^+_{/j}$ ($j \geq 1$) have the same
meaning as in the dual alphabet of section~\ref{sec:dual}. The symbols
$\Downarrow_j$ ($0 \leq j \leq n$) indicate that both sequences have match
length 0 and that the next non-skipped position is $j$ nucleotides
further. The new symbol $*$ indicates that the nature of the nucleotide
does not matter, as we will explain below.

Figure~\ref{fig:skip_dual} shows the read from Figure~\ref{fig:dual}
represented in the skip-3 dual encoding. It is important to note several
differences with Figure~\ref{fig:dual}. The first is that the symbols
$\Downarrow_j$ ($0 \leq j \leq n$) are not always associated with double
mismatches. For instance, the symbol $\Downarrow_0$ on the right side of
the read corresponds to a mismatch for the $+$ sequence only. This happens
whenever the $+$ and the $-$ sequences are mismatched in the same interval
between non-skipped positions (the mismatches do not need to be on the
same nucleotide).

Also observe that the symbol $\Downarrow_1$ is followed by a $*$ to
indicate that this nucleotide has no influence on the presence of a seed,
regardless 
. The reason is that there
is a double mismatch at the $\Downarrow_1$ symbol so there can be no seed.
But since the next potential seed starts only at the next tick mark,
\textit{i.e.} one nucleotide further downstream,

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip_dual.pdf}
\caption{\textbf{Example of skip dual encoding.}
The read of Figure~\ref{fig:dual} is represented in the skip dual alphabet.
The vertical bars above the read indicate non-skipped positions. The
symbols $\downarrow_{/j}^-$ and $\downarrow_{/j}^+$ ($j \geq 1$) have the
same meaning as in the dual alphabet. The symbols $\Downarrow_j$ indicate
that both sequences have match length 0 and that the next non-skipped
position is located $j$ nucleotides downstream. Other features are as in
Figure~\ref{fig:dual}.}
\label{fig:skip_dual}
\end{figure}

As in the previous section, let $a$, $b$, $c$ and $d$ denote the
probabilities of a double match, a mismatch against $-$, a mismatch
against $+$ and a double mismatch, respectively. This also corresponds to
the probabilities of occurrence of the symbols $\square$,
$\downarrow_{/j}^+$, $\downarrow_{/j}^-$ ($j \geq 1$) and $\Downarrow_j$
($0 \leq j \leq n$), respectively. With this information, we compute the
weighted generating functions of interest in the sections below.

\subsection{Segments following $\Downarrow_j$ ($1 \leq j \leq n)$}

After a $\Downarrow_j$ terminator, the match counter for both sequences is
at zero and the next potential seed starts only $j \geq 1$ nucleotides
further downstream. This means that the next $j$ nucleotides are just
filler so it does not matter which sequence they match or mismatch.

There are only two scenarios to consider: if the end of the read is within
the next $j$ nucleotides, then

\subsection{Segments following $\Downarrow_0$}

Hello world!

\subsection{Transfer matrix}

@@

We illustrate the strategy delineated above Using the same settings as in
section~\ref{sec:illdual} ($k=50$, $p=0.01$ and $\gamma=19$), except that
we use skip-9 seeds instead of exact seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=0.47]{curves_skip.pdf}
\caption{\textbf{Off-target seeding probabilities (skip seeds).}
The surfaces show the probability that seeding is off target for skip-9
seeds of size $\gamma=19$. Read size and error rate are the same as in
Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$. The
divergence rate $\mu$, defined as the probability that a given duplicate
differs from the target at any given position. Note the factor 1000 in
the scale of the z-axis.}
\label{fig:curves_skip}
\end{figure}

@@

We illustrate the strategy delineated above 
using the same settings as in section~\ref{sec:illdual} ($k=50$, $p=0.01$
and $\gamma=19$), except that we replace exact seed by MEM seeds.

\begin{figure}[h]
\centering
\includegraphics[scale=.48]{curves_mem.pdf}
\caption{\textbf{Off-target seeding probabilities (MEM seeds).}
The surfaces show the probability that seeding is off target for MEM seeds
of indicated minimum size $\gamma$. Read size and error rate are the same
as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and $p=0.01$.
The divergence rate $\mu$ is the probability that a given duplicate
differs from the target at any given position. Note the factor 1000 in the
scale of the z-axis.}
\label{fig:curves_mem}
\end{figure}

@@

The value of $\beta$ has little importance if $p$ is small. Say that
$p=0.01$ and that a read generates a hit in the genome such that 33
nucleotides need to be aligned after seeding. If the hit is random there
is a 99.99\% chance that at least 15 are mismatches. The denominator of
expression (\ref{eq:bayes}) is approximately $1 +
4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
support for the hypothesis that the hit is random is overwhelming.
Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
fewer nucleotides are mismatches. In this case, the denominator is
approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
10^{-9}$, the support for the hypothesis that the hit is not random is
overwhelming. If $p$ is small, we do not need to worry about the value of
$\beta$, one can choose for instance $\beta=1/2$ so that the term
$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).

Since the probability that the best candidate is a random sequence is
either very small or very large, it has no influence in the first case,
or it dominates the probability of a false positive in the second case.
For every read, we can use as confidence score the maximum of this
probability and of the estimated probability that seeding is off target.
